Relational Parametricity for Algebraically Indexed Types

### 2012-06-12 paper/geometry-twocolumn.tex
Major changes
- Refined set of contributions
- "Complete" the motivating examples section
- Fill in some references for the background section
- "Complete" the general framework section
- Sketch the "consequences" section
- Sketch the "discussion" section

We motivate this paper's goal of investigating algebraically-indexed
types and their relational interpretations by presenting several
examples. Our main set of examples comprise a novel type system for
writing programs that manipulate geometric data, while maintaining
geometric invariants through the types. \fixme{and the information
  flow examples}

@@

We now sketch the interpretation of the types $\tyPrim{vec}{e}$ and
$\forall t\mathord:\SynTransl_2.A$. In our calculus, each type has two
interpretations: an index-erasure interpretation that ignores the
indexing expression $e$, and a relational interpretation as a binary
relation on the index-erasure interpretation. For now, we assume we
are given some way of turning an expression $e$ of sort $\SynTransl_2$
into an actual element $\sem{e}\rho$ of the group $\Transl_2$, under
some environment $\rho$ which tells how to interpret the free
variables. The (simplified) index-erasure and relation interpretations
of the types $\tyPrim{vec}{e}$ are:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2 \\
    \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{t}\rho} = \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{e}\rho \}
  \end{array}
\end{displaymath}
The index-erasure interpetation of universal quantification $\forall
t\mathord:\SynTransl_2.A$ is simply the interpretation of the
underlying type $A$. The relational interpretation intersects the
relational interpretations of the type $A$ under all extensions of the
relational environment $\rho$:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\forall t\mathord:\SynTransl_2.A} & \tySem{A} \\
    \rsem{\forall t\mathord:\SynTransl_2.A}{}\rho & \bigcap\{ \rsem{A}{}{\rho'} \sepbar \rho' \in \extends{\rho}{i\mathord:\SynTransl_2} \}
  \end{array}
\end{displaymath}
The exact definition of $\extends{\rho}{i\mathord:\SynTransl_2}$
depends on which (possibly non-compositional) interpretation of
relational environments that we take. For now, it will suffice to
think of relational environments as simply mappings from free
variables to the group $\Transl_2$, and extension as just adding an
extra component to the tuple. The general index-erasure semantics will
be given in \autoref{sec:index-erasure-semantics}. The full relational
interpretation is in general more complex, due to the possibility of
non-compositional interpretations of free index variables. The
relational interpretation will be presented in
\autoref{sec:relational-semantics}.

Equipped with quantification over the group $\SynTransl_2$ of changes
of origin, and the indexed types $\tyPrim{vec}{\cdot}$, we can now
express origin invariance properties as types. In the previous
section, we discussed functions $f : \mathbb{R}^2 \times \mathbb{R}^2
\to \mathbb{R}^2$ that have the property that they preserve all
changes of origin. We can now express this property as a type:
\begin{displaymath}
  \mathrm{f} : \forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \times \tyPrim{vec}{t} \to \tyPrim{vec}{t}
\end{displaymath}
Spelling out the relational interpretation of this type using the
definitions above (and the standard relational interpretations for
tuple and function types), we obtain
\statementref{eq:f-invariant-frame}.

@@

\fixme{Need to make clear that the point is to make use of the
  algebraic indexing structure}

@@

However, it is possible to give vector addition a more precise type
that describes its effect on the chosen origin. 
\begin{displaymath}
  (+) : \forall t_1, t_2 \mathord: \SynTransl_2.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2}
\end{displaymath}
We can also negate vectors, yielding a vector which points in the
opposite direction. Negation negates translation arguments:
\begin{displaymath}
  \mathrm{negate} : \forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{-t}
\end{displaymath}
With the primitive operations of addition and negation of vectors, we
can define the derived operation of subtraction:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall t_1,t_2 \mathord:\SynTransl_2.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 - t_2} \\
    (-)\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
  \end{array}
\end{displaymath}

Given two points that have been constrained to be invariant with
respect to the same change of origin---i.e.~two values of type
$\tyPrim{vec}{t}$---we can compute their offset, which is a vector
expressed with respect to the null change of origin. The offset
operation is a special case of vector subtraction:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{offset} : \forall t \mathord:\SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{0} \\
    \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
  \end{array}
\end{displaymath}
The type of the vector addition operation can be specialised to the
case of moving a point by a vector:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{moveBy} : \forall t \mathord:\SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{0} \to \tyPrim{vec}{t} \\
    \mathrm{moveBy}\ [t]\ p\ v = p + v
  \end{array}
\end{displaymath}
The types we assign to the remaining vector space primitives, the zero
vector and multiplication by a scalar, do not describe any interesting
effect on translations:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}:\hspace{0.5em}}l}
    \mathrm{0} & \tyPrim{vec}{0}\\
    (*) & \tyPrimNm{real} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}
  \end{array}
\end{displaymath}

@@

\begin{example}
  The vector space operators and the properties that follow from their
  types allow us to establish a useful type isomorphism. Consider
  functions with types following the schema:
  \begin{displaymath}
    \tau_{n} \isDefinedAs \forall t\mathord:\SynTransl_2.\ \underbrace{\tyPrim{vec}{t} \to ... \tyPrim{vec}{t}}_{n+1\textrm{ times}} \to \tyPrimNm{real}
  \end{displaymath}
  Just by looking at the types $\tau_{n}$, we know that their
  inhabitants will be invariant under change of origin, due to the
  quantification over all $t$ in $\SynTransl_2$. Therefore, we may as well pick
  one of the input points as the origin and assume that all the other
  points are defined with respect to this chosen origin. This is
  similar to the common mathematical step of stating that, ``without
  loss of generality'', we may pick some point in a description of a
  problem to be the origin, as long as the problem statement is
  invariant under translation.

@@

\begin{example}
  An inexpressible type
  \begin{displaymath}
    \forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}
  \end{displaymath}
  FIXME: explain that it is possible to show that this type is
  uninhabited with the given operations.
\end{example}

@@

From basic linear algebra, we know that the vector operations of
addition, negation and scaling, while not invariant under change of
origin, are invariant under change of basis. Following the same
methodology as we did for origin invariance, we can express basis
invariance by means of preservation of relations indexed by changes of
basis. Change of basis is accomplished by applying an invertible
linear map, and the collection of all invertible linear maps on
$\mathbb{R}^2$ forms the group $\GL_2$.

\fixme{new relational interpretation of vectors}

\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e_B,e_t}} & \mathbb{R}^2 \\
    \rsem{\tyPrim{vec}{e_B,e_t}}{}\rho & \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \sem{e_B}\rho\vec{v} + \sem{e_t}\rho \}
  \end{array}
\end{displaymath}

\paragraph{Affine Geometry} The combination of a change of basis
transformation and a change of origin transformation is known as an
\emph{affine transformation}. A way of describing affine geometry is
as the collection of properties that are invariant under invertible
affine transformations.

\begin{eqnarray*}
  \mathrm{affComb} & : &
  \begin{array}[t]{@{}l}
    \forall B \mathord: \SynGL_2, t \mathord: \SynTransl_2.\\
    \hspace{0.2cm} \tyPrim{vec}{B,t} \to \tyPrimNm{real} \to \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,t}
  \end{array}
  \\
  (+) & : &
  \begin{array}[t]{@{}l}
    \forall B \mathord: \SynGL_2, t_1,t_2 \mathord: \SynTransl_2. \\
    \hspace{0.2cm}\tyPrim{vec}{B,t_1} \to \tyPrim{vec}{B,t_2} \to \tyPrim{vec}{B,t_1 + t_2}
  \end{array}
  \\
  \mathrm{negate} & : & \forall B \mathord: \SynGL_2, t \mathord: \SynTransl_2.\ \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,-t} \\
  0 & : & \forall B \mathord: \SynGL_2.\ \tyPrim{vec}{B,0} \\
  (*) & : & \forall B \mathord: \SynGL_2.\ \tyPrimNm{real} \to \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0}
\end{eqnarray*}

\fixme{we now have all the operations of affine geometry, along with
  the property that they are invariant under affine transformations.}

\paragraph{Euclidean Geometry} Euclidean geometry extends affine
geometry with the operation of \emph{dot product} (also known as the
\emph{inner product}) of two vectors, which produces a real
number. The standard Euclidean dot product is defined on the
coordinate representation in the following way:
\begin{displaymath}
  (x_1,y_1) \cdot (x_2,y_2) = x_1x_2 + y_1y_2
\end{displaymath}
% The dot product is used to define properties of vectors such as their
% \emph{norm} $||\vec{v}|| = \vec{v}\cdot\vec{v}$, which is the square
% of the length of the vector $\vec{v}$, and the notion of
% orthogonality: two vectors $\vec{v_1}$ and $\vec{v_2}$ are orthogonal
% if $\vec{v_1}\cdot\vec{v_2} = 0$.

To incorporate the operation of dot product into our calculus, we must
give it a type. It is not the case that the dot product is invariant
under the action of either of the groups $\GL_2$ or $\Transl_2$ that
we have considered so far. In fact, the dot product is invariant under
the subgroup $\Orth_2$ of $\GL_2$ of \emph{orthogonal} linear
transformations. The group $\Orth_2$ consists of invertible linear
maps whose determinant is either $1$ or $-1$. We introduce orthogonal
transformations as a new sort $\SynOrth_2$, and overload the
multiplicative group operations syntax for inhabitants of
$\SynOrth_2$. We also assume a function $\iota_O$ that takes $e :
\SynOrth_2$ to $\iota_O(e) : \SynGL_2$. We may now assign the following type
to the dot product:
\begin{displaymath}
  (\cdot) : \forall O \mathord: \SynOrth(2).\ \tyPrim{vec}{\iota_OO, 0} \to \tyPrim{vec}{\iota_OO, 0} \to \tyPrimNm{real}
\end{displaymath}

@@

The cross product of two vectors is defined on coordinate
representations as follows:
\begin{displaymath}
  (x_1,y_1) \times (x_2,y_2) = x_1y_2 - x_2y_1
\end{displaymath}
Geometrically, the cross product is the signed area of the
parallelogram described by the pair of input vectors. It turns out
that the value of the cross product of two vectors varies with the
determinant of the matrix describing the basis of the vector space. We
therefore augment our calculus with a new sort $\SynGL_1$ of scale
factors. Semantically, $\SynGL_1$ ranges over the non-zero real
numbers and forms an abelian group, which we write
multiplicatively. We also add two new operations: determinant, $\det
B$, which takes an inhabitant of $\SynGL_2$ to its determinant in
$\SynGL_1$, and the operation of absolute value that takes scaling
factors to scaling factors which we write as $|e|$. We also refine the
type $\tyPrimNm{real}$ of real numbers to take a parameter of sort
$\SynGL_1$: $\tyPrim{real}{e}$, where the old type $\tyPrimNm{real}$
is just $\tyPrim{real}{1}$. The collection of operations on real
numbers with scaling factors is shown in \autoref{fig:real-ops}.

With the additional sort $\SynGL_1$, and the refined type of real
numbers, we can state the type of the cross product.
\begin{displaymath}
  (\times) : \forall B \mathord: \SynGL_2.\ \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0} \to \tyPrim{real}{\det B}
\end{displaymath}
For orthogonal transformations $O : \Orth_2$, the absolute value of
the determinant is always $1$, so we assume the equation $|\det
(\iota_O O)| = 1$ to hold.

\begin{example}
  Using the structure we have defined in this subsection, we can
  define the following function that computes the area of a
  triangle. By the type of this function, we can see that the area of
  a triangle is invariant under orthogonal transformations and
  arbitrary translations \fixme{i.e. isometries}.
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{area} : \forall O\mathord:\SynOrth_2, t\mathord:\SynTransl_2.\ \\
      \hspace{0.8cm}\tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{real}{1} \\
      \mathrm{area}\ [O]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} * \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
    \end{array}
  \end{displaymath}
  In the body of this function, we have performed the calculation in
  several steps, each of which removes some of the symmetry described
  by the type of $\mathrm{area}$. First we have computed the two
  offset vectors $p_2 - p_1$ and $p_3 - p_1$. These operations remove
  the effect of translations on the result, in exactly the same way as
  the type isomorphism we demonstrated above \fixme{ref}. We then
  compute the cross product of the two vectors to determine the area
  of the parallelogram described by the sides of the triangle. This
  value has type $\tyPrim{real}{\det (\iota_O O)}$. We know that the
  determinant of an orthogonal transformation is always either $-1$ or
  $1$, so we have a result which varies according to possible
  reflections of the space. To remove this variation, we apply the
  $\mathrm{abs}$ function to leave us with a value of type
  $\tyPrim{real}{|\det (\iota_O O)|}$. This type is equal to the type
  $\tyPrim{real}{1}$. Finally, we multiply by $\frac{1}{2}$ to recover
  the area of the triangle, rather than the whole parallelogram.
\end{example}

% \begin{enumerate}
% \item Basis invariance
% \item Orthogonal transform invariance for dot product
% \item Rotational invariance (the type of the cross-product)
% \item Scale invariance (generalisation of units of measure)
% \end{enumerate}

% In some cases, we can use the invariance as a programming tool
% (e.g. type isomorphisms), but at others we can only use it from the
% outside.

@@

\subsection{Scale Invariance and Dimension Analysis}
\label{sec:scale-invariance}

\begin{enumerate}
\item Explicitly link scale invariance to units of measure system
\end{enumerate}

Need a function $\iota_1 : \SynGL_1 \to \SynGL_2$. Such that $\det (\iota_1 s) = s^2$

\subsection{Information Flow}
\label{sec:information-flow}

Blah blah blah. Need to work this out in terms of representation indepdence.

\begin{enumerate}
\item Think of information flow as passing around unforgable tokens
  \begin{enumerate}
  \item Give an example (given a piece of data protected by a token
    and a token, can we extract the data?)
  \item Need to make sure that tokens are unforgable
  \end{enumerate}
\item Could either enforce unforgability of tokens by some dynamic
  means, or by static type-based means
  \begin{enumerate}
  \item Leave the dynamic method as vague
  \item Ingredients of the static method: run-time tokens are just
    dummy values ($*$), the token type is parameterised by logical
    formulae, and the relational meaning is derived from an assignment
    of boolean values to atoms.
  \item Explain this as invariance under changes of satisfaction of
    the participants. Relate to descriptions of logics as rules that
    don't care about truth, only its preservation.
  \end{enumerate}
\item Cite Sabelfeld and Sands: ``A PER Model of Secure Information
  Flow in Sequential Programs''. And also Tse and Zdancewic.
\item Go a step further by considering equivalence relations on
  word32, and their lattice structure. Link this to quantative
  information flow. Will also need to include singleton types.
\end{enumerate}

\subsection{Basic Continuity Analysis}
\label{sec:continuity-analysis}

Doing epsilon-delta proofs!

Have a new sort $\mathsf{R}^{>0}$, representing positive distances
between real numbers.

\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{real}{e}} & \mathbb{R} \\
    \rsem{\tyPrim{real}{e}}{}\rho & \{ (x,x') \sepbar |x-x'| < \sem{e}\rho \}
  \end{array}
\end{displaymath}

>>>>

We motivate this paper's goal of investigating algebraically-indexed
types and their relational interpretations by presenting several
examples. Our leading example is a novel type system for writing
programs that manipulate geometric data, while maintaining geometric
invariants through the types. The other examples are enforcement of
information flow properties via indexing by a logical theory
(\autoref{sec:information-flow}), and incorporation of metric
properties into types (\autoref{sec:continuity-analysis}).

@@

We now sketch the interpretation of the types $\tyPrim{vec}{e}$ and
$\forall t\mathord:\SynTransl_2.A$. In our calculus, each type has two
interpretations: an index-erasure interpretation that ignores the
indexing expression $e$, and a relational interpretation as a binary
relation on the index-erasure interpretation. For now, we assume we
are given some way of turning an expression $e$ of sort $\SynTransl_2$
into an actual element $\sem{e}\rho$ of the group $\Transl_2$, under
some environment $\rho$ which tells how to interpret the free
variables. The (simplified) index-erasure and relation interpretations
of the types $\tyPrim{vec}{e}$ are:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2 \\
    \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{e}\rho} = \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{e}\rho \}
  \end{array}
\end{displaymath}
The index-erasure interpetation of universal quantification $\forall
t\mathord:\SynTransl_2.A$ is simply the interpretation of the
underlying type $A$. The relational interpretation intersects the
relational interpretations of the type $A$ under all extensions of the
relational environment $\rho$:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\forall t\mathord:\SynTransl_2.A} & \tySem{A} \\
    \rsem{\forall t\mathord:\SynTransl_2.A}{}\rho & \bigcap\{ \rsem{A}{}{\rho'} \sepbar \rho' \in \extends{\rho}{t\mathord:\SynTransl_2} \}
  \end{array}
\end{displaymath}
The exact definition of $\extends{\rho}{t\mathord:\SynTransl_2}$
depends on which (possibly non-compositional) interpretation of
relational environments that we take. For now, it will suffice to
think of relational environments as simply mappings from free
variables of sort $\SynTransl_2$ to the group $\Transl_2$, and
extension as just adding an extra component to the tuple. The general
index-erasure semantics will be given in
\autoref{sec:index-erasure-semantics}. The relational interpretation
is in general more complex, due to the possibility of
non-compositional interpretations of free index variables. The
relational interpretation will be presented in
\autoref{sec:relational-semantics}.

In the previous section, we discussed functions $f : \mathbb{R}^2
\times \mathbb{R}^2 \to \mathbb{R}^2$ that have the property that they
preserve all changes of origin. We can now express this property as a
type:
\begin{displaymath}
  \mathrm{f} : \forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \times \tyPrim{vec}{t} \to \tyPrim{vec}{t}
\end{displaymath}
Spelling out the relational interpretation of this type using the
definitions above (and the standard relational interpretations for
tuple and function types), we obtain
\statementref{eq:f-invariant-frame}.

@@

@@

However, it is possible to give vector addition a more precise type
that shows how it is variant with change of origin:
\begin{displaymath}
  (+) : \forall t_1, t_2 \mathord: \SynTransl_2.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2}
\end{displaymath}
We can also negate vectors, yielding a vector which points in the
opposite direction. Negation negates translation arguments:
\begin{displaymath}
  \mathrm{negate} : \forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{-t}
\end{displaymath}
With the primitive operations of addition and negation of vectors, we
can define the derived operation of subtraction:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall t_1,t_2 \mathord:\SynTransl_2.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 - t_2} \\
    (-)\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
  \end{array}
\end{displaymath}

Given two points that have been constrained to be invariant with
respect to the same change of origin---i.e.~two values of type
$\tyPrim{vec}{t}$---we can compute their offset, which is a vector
expressed with respect to the null change of origin. The offset
operation is a special case of vector subtraction:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{offset} : \forall t \mathord:\SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{0} \\
    \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
  \end{array}
\end{displaymath}
Note how the type of the result of $\mathrm{offset}$ simplifies from
$\tyPrim{vec}{t-t}$ to $\tyPrim{vec}{0}$. The algebraic structure on
the indexing theory induces type equalities.  The type of the vector
addition operation can be specialised to the case of moving a point by
a vector:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{moveBy} : \forall t \mathord:\SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{0} \to \tyPrim{vec}{t} \\
    \mathrm{moveBy}\ [t]\ p\ v = p + v
  \end{array}
\end{displaymath}
The types we assign to the remaining vector space primitives, the zero
vector and multiplication by a scalar, do not describe any interesting
effect on translations:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}:\hspace{0.5em}}l@{\hspace{4em}}l@{\hspace{0.5em}:\hspace{0.5em}}l}
    \mathrm{0} & \tyPrim{vec}{0}&
    (*) & \tyPrimNm{real} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}
  \end{array}
\end{displaymath}

@@

\begin{example}\label{ex:type-iso}
  The vector space operators and the properties that follow from their
  types allow us to establish a useful type isomorphism. Consider
  functions with types following the schema:
  \begin{displaymath}
    \tau_{n} \isDefinedAs \forall t\mathord:\SynTransl_2.\ \underbrace{\tyPrim{vec}{t} \to ... \tyPrim{vec}{t}}_{n+1\textrm{ times}} \to \tyPrimNm{real}
  \end{displaymath}
  Just by looking at the types $\tau_{n}$, we know that their
  inhabitants will be invariant under change of origin, due to the
  quantification over all $t$ in $\SynTransl_2$. Therefore, we may as well pick
  one of the input points as the origin and assume that all the other
  points are defined with respect to this chosen origin. This is
  similar to the common mathematical step of stating that, ``without
  loss of generality'', we may pick some point in a description of a
  problem to be the origin, as long as the problem statement is
  invariant under translation.

@@

\begin{example}\label{ex:uninhabited-type}
  To this point we have emphasised the properties that we can deduce
  about programs from their types, so called ``free theorems''. By
  refining our relational interpretations of types appropriately it is
  also possible to show that certain types are uninhabited. For
  example, the type $\forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t
    + t} \to \tyPrim{vec}{t}$ does not have any inhabitants. We
  demonstrate this in \autoref{sec:types-indexed-abelian-groups}. In
  order to obtain such a result, we require a more sophisticated
  relational interpretation of free indexing variables than the one we
  sketched above.
\end{example}

@@

From basic linear algebra, we know that the vector operations of
addition, negation and scaling, while not invariant under change of
origin, are invariant under change of basis. Following the same
methodology as we did for origin invariance, we can express basis
invariance by means of preservation of relations indexed by changes of
basis. Change of basis is accomplished by applying an invertible
linear map, and the collection of all invertible linear maps on
$\mathbb{R}^2$ forms the group $\GL_2$. We reflect in our language the
group $\GL_2$ as a new indexing sort $\SynGL_2$, which has
(non-abelian) group structure that we will write multiplicatively. We
also extend the $\tyPrimNm{vec}$ type to be indexed by elements of
$\SynGL_2$ as well as $\SynTransl_2$. Thus, $\tyPrim{vec}{B,t}$ is a
vector which varies with change of basis $B$ and change of origin
$t$. Formally, the index-erasure and (simplified) relational semantics
of this type are as follows:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e_B,e_t}} & \mathbb{R}^2 \\
    \rsem{\tyPrim{vec}{e_B,e_t}}{}\rho & \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = (\sem{e_B}\rho)\vec{v} + \sem{e_t}\rho \}
  \end{array}
\end{displaymath}

\paragraph{Affine Geometry} The combination of an invertible linear
map and a translation is an \emph{affine transformation}. We can now
assign types to all the primitive affine and vector space operations
that indicates how they they behave with respect to affine
transformations:
\begin{eqnarray*}
  \mathrm{affComb} & : &
  \begin{array}[t]{@{}l}
    \forall B \mathord: \SynGL_2, t \mathord: \SynTransl_2.\\
    \hspace{0.2cm} \tyPrim{vec}{B,t} \to \tyPrimNm{real} \to \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,t}
  \end{array}
  \\
  (+) & : &
  \begin{array}[t]{@{}l}
    \forall B \mathord: \SynGL_2, t_1,t_2 \mathord: \SynTransl_2. \\
    \hspace{0.2cm}\tyPrim{vec}{B,t_1} \to \tyPrim{vec}{B,t_2} \to \tyPrim{vec}{B,t_1 + t_2}
  \end{array}
  \\
  \mathrm{negate} & : & \forall B \mathord: \SynGL_2, t \mathord: \SynTransl_2.\ \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,-t} \\
  0 & : & \forall B \mathord: \SynGL_2.\ \tyPrim{vec}{B,0} \\
  (*) & : & \forall B \mathord: \SynGL_2.\ \tyPrimNm{real} \to \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0}
\end{eqnarray*}

\paragraph{Euclidean Geometry} Euclidean geometry extends affine
geometry with the operation of \emph{dot product} (also known as the
\emph{inner product}) of two vectors. The standard Euclidean dot
product is defined on the coordinate representation in the following
way:
\begin{displaymath}
  (x_1,y_1) \cdot (x_2,y_2) = x_1x_2 + y_1y_2
\end{displaymath}
% The dot product is used to define properties of vectors such as their
% \emph{norm} $||\vec{v}|| = \vec{v}\cdot\vec{v}$, which is the square
% of the length of the vector $\vec{v}$, and the notion of
% orthogonality: two vectors $\vec{v_1}$ and $\vec{v_2}$ are orthogonal
% if $\vec{v_1}\cdot\vec{v_2} = 0$.
To incorporate the operation of dot product into our calculus, we must
assign it a type. The dot product is not invariant under the action of
either of the groups $\GL_2$ and $\Transl_2$ that we have considered
so far. However, the dot product is invariant under the subgroup
$\Orth_2$ of $\GL_2$ of \emph{orthogonal} linear transformations:
invertible linear maps whose determinant is either $1$ or $-1$. We
introduce orthogonal transformations as a new sort $\SynOrth_2$, and
overload the multiplicative group operations syntax for inhabitants of
$\SynOrth_2$. We also assume a function $\iota_O$ that takes $e :
\SynOrth_2$ to $\iota_O(e) : \SynGL_2$. We may now assign a type to
the dot product, using quantification over $\SynOrth_2$:
\begin{displaymath}
  (\cdot) : \forall O \mathord: \SynOrth_2.\ \tyPrim{vec}{\iota_OO, 0} \to \tyPrim{vec}{\iota_OO, 0} \to \tyPrimNm{real}
\end{displaymath}

@@

The cross product of two vectors is defined on coordinate
representations as $(x_1,y_1) \times (x_2,y_2) = x_1y_2 - x_2y_1$.
Geometrically, the cross product is the signed area of the
parallelogram described by the pair of input vectors. Under change of
basis by an invertible linear transformation $B$, the value of the
cross product of two vectors varies with the determinant of $B$. This
corresponds to the scaling of the plane by the change of basis
transformation.

We therefore augment our calculus with a new sort $\SynGL_1$ of scale
factors (i.e.~$1$-dimensional invertible linear maps). Semantically,
$\SynGL_1$ ranges over the non-zero real numbers and forms an abelian
group, which we write multiplicatively. We also add two new
operations: determinant, $\det B$, which takes an inhabitant of
$\SynGL_2$ to its determinant in $\SynGL_1$, and the operation of
absolute value that takes scaling factors to scaling factors which we
write as $|e|$. We also refine the type $\tyPrimNm{real}$ of real
numbers to be indexed by the sort $\SynGL_1$: $\tyPrim{real}{e}$,
where the old type $\tyPrimNm{real}$ is just $\tyPrim{real}{1}$. The
collection of operations on real numbers indexed by scaling factors is
shown in \autoref{fig:real-ops}.

With the additional sort $\SynGL_1$, and the refined type of real
numbers, we can assign a type to the cross product:
\begin{displaymath}
  (\times) : \forall B \mathord: \SynGL_2.\ \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0} \to \tyPrim{real}{\det B}
\end{displaymath}
Since the absolute value of the determinant of an orthogonal
transformation is always $1$, we assume $|\det (\iota_O O)| = 1$ to
hold.

\begin{example}\label{ex:area-of-triangle-1}
  Using the operations we have defined in this subsection, we can
  compute the area of a triangle. 
% By the type of this function, we can
%   see that the area of a triangle is invariant under orthogonal
%   transformations and arbitrary translations. Such transformations are
%   known as isometries: distance preserving maps.
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{area} : \forall B\mathord:\SynGL_2, t\mathord:\SynTransl_2.\ \\
      \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{real}{|\det B|} \\
      \mathrm{area}\ [B]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} * \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
    \end{array}
  \end{displaymath}
  The calculation is performed in several steps, each of which removes
  some of the symmetry described by the type of
  $\mathrm{area}$. First, the two offset vectors $p_2 - p_1$ and $p_3
  - p_1$ are computed. These operations remove the effect of
  translations on the result, in exactly the same way as the type
  isomorphism we introduced in \exref{ex:type-iso}. We then compute
  the cross product of the two vectors to determine the area of the
  parallelogram described by the sides of the triangle. This value has
  type $\tyPrim{real}{\det B}$, and we have removed some of the
  symmetry due to invertible linear maps. This value still varies with
  the sign of the determinant, so we remove this symmetry by applying
  the $\mathrm{abs}$ function to leave us with a value of type
  $\tyPrim{real}{|\det B|}$. Finally, we multiply by the constant
  $\frac{1}{2}$ to recover the area of the triangle, rather than the
  whole parallelogram.

  If we specialise $\mathrm{area}$ to just orthogonal transformations,
  we obtain the following type, as a consequence of the equation
  $|\det (\iota_OO)| = 1$:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{area} : \forall O\mathord:\SynOrth_2, t\mathord:\SynTransl_2.\ \\
      \hspace{0.8cm}\tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{real}{1}
    \end{array}
  \end{displaymath}
  This type demonstrates that the area of a triangle is invariant
  under orthogonal transformations and translations. Such combined
  transformations are known as isometries---distance preserving maps.
\end{example}

@@

\subsection{Scale Invariance and Dimensional Analysis}
\label{sec:scale-invariance}

Indexing types by scaling factors brings us to the original
inspiration for the current work: Kennedy's interpretation of his
units of measure type system via scaling invariance
\cite{kennedy97relational}. Kennedy demonstrates how interpreting
types in terms of scaling invariance brings the techniques of
dimensional analysis \fixme{ref} to bear on programming. The types we
have assigned to the real number arithmetic operations in
\autoref{fig:real-ops} are exactly the types Kennedy assigns in his
units of measure system, except for the absolute value operation. The
effect of our assigned type is that, semantically, we are indexing by
non-zero scaling factors, where Kennedy indexes by strictly positive
scaling factors. We further examine the typing of the absolute value
operation in \autoref{sec:types-indexed-abelian-groups}.

In our two-dimensional setting, we can make additions to Kennedy's
one-dimensional scaling invariance. We add an operation $\iota_1$
that, semantically, takes scale factors in $\GL_1$ to invertible
linear maps in $\GL_2$: i.e.~takes numbers $s$ to matrices $\left(
  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$.  This
operation satisfies the equation $\det (\iota_1 s) = s^2$, indicating
that if we scale the plane by $s$ in all directions, we scale areas by
$s^2$. This is illustrated by the next example.

\begin{example}\label{ex:area-of-triangle-2}
  Just as we specialised the type of $\mathrm{area}$ function to
  orthogonal transformations in \exref{ex:area-of-triangle-1}, we can
  also specialise $\mathrm{area}$'s type to scaling
  transformations. This yields the type:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{area} : \forall s\mathord:\SynGL_1, t\mathord:\SynTransl_2.\ \\
      \hspace{0.8cm} \tyPrim{vec}{\iota_1s, t} \to \tyPrim{vec}{\iota_1s, t} \to \tyPrim{vec}{\iota_1s, t} \to \tyPrim{real}{s^2}
    \end{array}
  \end{displaymath}
  As expected, the area of a triangle varies with the square of
  scalings of the plane, and this is reflected in the type.
\end{example}

Purely scaling linear maps, as generated by $\iota_1$, commute with
all other invertible linear maps, and form the centre of the group
$\GL_2$. Thus we add the equation $\iota_1(s)B = B\iota_1(s)$ to our
calculus.
